Variant calling with transformers

### 2022-06-08 manuscript/manuscript.tex
manuscript updates

@@

Models were implemented in Pytorch (v1.10, Paszke et al 2019). Models were trained for 25 epochs unless noted otherwise, using the Adam optimizer and a learning rate of $5e-5$.

We use a standard cross-entropy loss function to compare predicted haplotypes to known haplotype sequences for each region, and simply compare the prediction at window index (position) $i$ to the true base at index $i$, where $i \in 1..g$.  While we obtain satisfactory results with this approach, two issues were apparent. First, predicted haplotype sequences may have inserted or deleted bases relative to the true haplotypes. Because incorrect insertions and deletions cause mismatches at every downstream (higher index) position, such mistakes are penalized more heavily when they occur at lower indexes than higher indexes. Thus the model learns to generate high quality results for the first portion of the window where inaccuracies are punished severely, and is less concerned with accuracy at high indexes.

>>>>

\begin{figure}[htp]
	\includegraphics[scale=0.5]{./jovian_architecture_diagram.png}
	\caption{ Overview of network architecture, numbers in parentheses indicate size of tensors  }
	\label{fig:architecture}
\end{figure}

@@

Models were implemented in Pytorch (v1.10, Paszke et al 2019). Models were trained for 25 epochs unless noted otherwise, using the Adam optimizer and a learning rate of $5e-5$ and batch size of 512.

For a given haplotype we use a standard cross-entropy loss function to compare the prediction to the target sequence. However, two haplotypes exist for each region and they have no inherent order. For instance, a single heterozygous variant can appear on either haplotype, and the model cannot learn predict which because the assigment is random. To make the predictions opaque to haplotype order, we compute the loss in both haplotype configurations - predicted haplotype 0 with target haplotype 0 and 1 with 1, as well as predicted haplotype 1 with target haplotype 0, and 0 with 1, and backpropogate gradients only for the configuration with the lowest loss.
